Writing Trash Bin

### 2024-01-23 paper.qmd
started doing summary of general ideas (no fine details)

# 1. Introduction
Think back to the last time you made some sort of data visualisation. What was the purpose of that visualisation? Was it to better understand your data? Was it to help you make a decision? Was it to to communicate that decision to someone else? Now think about the last time you expressed some form of uncertainty. Was it a set of numerical confidence intervals? Maybe they were expressed as a set of values in a table. Did you consider visualising your uncertainty instead? There are many stages in our analysis that benefit from the power of data visualisation, however this does not mean it is always done with success. Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set. The importance of data driven discovery is highlighted by data sets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg]. Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information when visualised. Anscombe quartet is shown in @fig-anscombe, because describing the data is never the same as seeing it. Instead of having to repeatedly check endless hypothesis to find interesting numerical features, visualisations **tell** us what is important about our data. This powerful aspect of data visualisation is poorly or seldom used in later stages when we are communicating our findings, specifically with respect to uncertainty.

@@

- industrial accidents have a "tollerable" risk chart based on risk and number of fatalities @Spiegelhalter2017
# 2. Defining Uncertainty
## 2.1 Coloquially
## 2.2 Through Taxonomies
Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature. @Lipshitz1997 even commented that “there are almost as many definitions of uncertainty as there are treatments of the subject”. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is. 
- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016

Papers that mention uncertainty often use a single sentance definition that suits the purpose of the paper, while papers that seek to define uncertainty use a taxonomy. While

More commonly uncertainty is defined using a taxonomy rather than a strict definition.

@@

*Notes from other section*
- The usually uncertainty is only considered at the final stage but it needs to be considered through all stages. The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
- @Meng2021 highlights that data needs to be treated as an output from another process rahter than an value neutral input.
- If the distinction between noise and uncertainty is largely task dependent, considerations of quantifying and expressing uncertainty should be considered for the entire duraction of a project; another concept that has been independently identified by several authors [@Kinkeldey2014; @Refsgaard2007]
- The most encompassing definition of uncertainty I have seen comes from @utypo who define uncertainty as **"any deviation from the unachievable ideal of completely deterministic knowledge of the relevant system"**. This definition encapsulates many concepts adjacent to randomness such as predictions, probability mass distributions (PMF), estimate error, and any data that is not a set of deterministic outcomes. This is the definition of uncertainty I will use for the rest of this report.
- While quantification of the uncertainty that arrises from statistical philosophy and data provenance are beyond the scope of this paper, how this uncertainty is communicated is not. The ontological process discussed by @Otsuka2023 where information is tossed as we boil our information down into statistical objects requires us to map our statistical objects back to real world entities. This means the noise introduced and signal lost in earlier stages of the data analysis need to be identified, quantified and expressed in terms of real world outcomes. Visualisation is often the final stage of larger data science project and the down stream effects of poor abstraction and problem at the lower levels become impossible to ignore [@Munzner2009]. 
- Approaches to communicating limited understanding include 1) explicit model uncertainty 2) Qualitative scales expressed as strength of evidence (e.g. high to low certainty of evidence) 3) Acknowledge limitations and their possible impact 4) Acknowledged ignorance @Spiegelhalter2017

@@

- @Spiegelhalter2017 Establishes that risk is known probabilities and uncertainty is estimated probabilities (i.e. more real world). Suggest uncertainty is what we have when you cannot accurately quantify risk
- Highlights the lack of a clear definition of uncertainty "it depends on who you ask" @Spiegelhalter2017
- ambiguity is also poorly defined, as it can be basically the same as uncertainty (uncertain probabilities) or contested values @Spiegelhalter2017
- A lot of taxonomies for uncertainty, simple is "known knowns, known unknowns, and unknown unknowns", ka (known, unknown, nad unknowable), (risk, uncertainty, ignorance). This also breaks down into the framework of aleatory uncertainty (unavoidable randomness), epistemic uncertatiny (uncertainty about structure, still quantified), ontological uncertainty (uncertain about entire modelling process). @Spiegelhalter2017
- Can discuss each uncertainty in isolation using distributions, ranges, and lists of alternatives, but the total cumulative uncertainty from every stage is more difficult (near impossible) @Spiegelhalter2017
# 3.General Attitude Towards Uncertainty + Its communication
I dont think this is worth a section but it is mentioned a lot in .
- the role of unertainty visualisation is to communicate accurate information
- problem of communicating to a group thing (avoid communicating inaccurate information)
- how people feel about uncertainty is irrelevant
  - natural disasters make people uncomfortable but to include it in a conversation about whether or not we should express it is rediculous.
- Numerical risks: Use absolute risks (but relative risk when catastrophic event), for single unique event use percent chance, express change as a proportion, frequency, or percentage (and be clear about reference class), avoid framing bias by showing risk with and without the outcome, keep the denominator fixed when making comparisons, explicit about time interval, For more knowledgeable audiences consider providing epistemic uncertainty and confidence in analysis @Spiegelhalter2017

@@

- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
# 5 Great Examples
- best things you have seen and why they are the best
- can't have a best practices because you need experimental results and my point is that the experiments aren't testing the right things
- VSUP, the map interactive map that susuan showed me, the pixel map

>>>>

# 1. Introduction
From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming. So overwhelming in fact, that it is not uncommon for people to become so overwhelmed by the information they refuse to engage with it at all. In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable. While websites that summarise insurance plans or hotel options down into a clear table make a living off it, few tools are as powerful or effective as a visualisation. Information visualisations remain one of the most powerful tools for fast and reliable science communication. Effectively leveraging visualisations to make them more effective is therefore key in facilitating simple and clear communication of science and other ideas.

Think back to the last time you made some sort of data visualisation. What was the purpose of that visualisation? Was it to better understand your data? Was it to help you make a decision? Was it to to communicate that decision to someone else? Now think about the last time you expressed some form of uncertainty. Was it a set of numerical confidence intervals? Maybe they were expressed as a set of values in a table. Did you consider visualising your uncertainty instead? There are many stages in our analysis that benefit from the power of data visualisation, however this does not mean it is always done with success. Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set. The importance of data driven discovery is highlighted by data sets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg]. Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information when visualised. Anscombe quartet is shown in @fig-anscombe, because describing the data is never the same as seeing it. Instead of having to repeatedly check endless hypothesis to find interesting numerical features, visualisations **tell** us what is important about our data. This powerful aspect of data visualisation is poorly or seldom used in later stages when we are communicating our findings, specifically with respect to uncertainty.

@@

- industrial accidents have a "tollerable" risk chart based on risk and number of fatalities @Spiegelhalter2017

# 2. Defining Uncertainty
Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature. @Spiegelhalter2017 even commented that the definition of uncertainty essentially "depends on who you ask". This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is. In this section we will establish exactly what we mean when we talk about uncertainty.

## 2.1 Coloquial Definitions
The most common definition of uncertainty used in papers that discuss visualising it are colloquial. Most definitions make some nod towards variance, precision, trust, error, missing values or another related concepts. Since *uncertainty*, unlike *probability*, does not have a strict definition that is established in mathematics, there is large variety in its usage. Some consider it to be synonymous with probabiltiy, some consider it to be an encompassing umbrella term of which probability is only an element, and some consider it to by something else entirely. This lack of a definition leads to uncertainty being swapped out freely with other vague words such as ambiguity and risk. Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". The issue with these broad definitions is is that in order to visualize something, it needs to be quantified, and in order to quantify something, we need to know exactly what it is we are counting. This discrepancy appears regularly in the uncertainty visualisation literature. Often a vague definition such as "uncertainty is anything that isn't deterministic", or even a refusal to define uncertainty at all in an experiments introduction will be accompanied by a highly specific quantification of uncertainty as a probability density distribution of a forecast when providing the visualisation. Ironically it seems that many of the papers that refer to themselves as "uncertainty" papers are just suffering from a form of uncertainty, a lack of precision.[^1]

This lack of precision has caused several pernicious issues in the literature, including different representation of uncertainty being compared despite not being realistic substitutes (with no acknowledgement of this from the authors); a lack of understanding as to *why* uncertainty should be included in a visualisation; a large swath of literature focused on a minute aspect of the entire uncertainty space with little attention to creative solutions to real world visualisation problems; and a swath of newly designed visualisations that are disconnected from the problems that they hope to solve. These problems means the literature either needs to transition away from the broad umbrella term "uncertainty", or consider the full range of sources of uncertainty and how they can be communicated when designing an experiment.[^2]

[^1]: Flesh out note - I think several of the sentences in this paragraph need examples (or quatified evidence) from the spreadsheet definition part
[^2]: I am not sure if I should mention this here because I go into depth in the literature review and these comments might come across as unjustified if they are left in this section. Maybe this is more of a conclusion paragraph.

## 2.2 Taxonomy Definitions
While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot. When authors try to make a considerable effort to define what we mean when we talk about uncertainty, they often express it in the form of a taxonomy.

When reading any literature about uncertainty or an adjacent topic you will be overrun with more ways to organise uncertainty than you will know what to do with. The problem is so pervasive that it is almost impossible for two randomly selected papers that both claim to be about uncertainty will use the same definition.

Some taxonomies contain overlapping terminology but also have alternative definitions and categories. For example *___* organises uncertainty into the two categories, Aleatroy (inherrent randomness) and epistemic (lack of knowledge), but *___* organises uncertainty into the three categories aleatory (inherrent randomness in a forecast) vs epistemic (uncertainty in structure and parameters of statistical models) vs ontological (uncertainty about the entire modelling process as a description of reality). Other taxonomies disucuss the degree of measurability

Other taxonomies conflate features of the data with features of the uncertainty:
- measurement (uncertainty on a single varible), spatial (uncertainty in spatial data), temporal (uncertainty in a time data)

there are even taxonomies

What is worse is that the reason for using one taxonomy over another, or why a new taxonomy was established in the first place is rarely discussed. This leaves new authors in the field with a wide array of choices and no information that would allow them to determine which is most effective for them. The complete mess of this literature makes it more enticing for authors to just define their own definition of uncertainty, which only adds to this problem. This is not done in ignorance of the issue either, many papers that comment on the unending list of uncertainty definitions will still provide a new taxonomy of their own design. I would almost find this situation comical if I were not one of the people forced to sort through this mess.

There are a handful of problems that is created by this lack of consensus. 
The first issues is that 
Which leaves the definition of aleatory and epistemic uncertainty confused.

 Below is a list of ways some papers have organised uncertainty:

- scientific methods (variance, distribution) vs human judgments (disagreement among parties)
- Random vs systematic uncertainty
- statistical (can provide a continuous pdf) vs bounded (know values will fall in a certain range) uncertainty

- accuracy vs precision
- Uncertainty from discovery, assessment and execution
- risk vs uncertainty
- Source of uncertainty (model structure, parameters, forcing functions, intial state, model operation) and diagnostic (past/current) vs prognostic (forecasting)
- uncertainty is organised according to Satience (legal moral, societal, institutional, proprietary, situational) and level of severity (high, medium, low)

There are several issues that become apparent when looking at this list. First of all, 
Looking at these taxonomies there are three key features of interest that become apparent. People seem to be interested in the:

1) What stage of the analyis or what group is causing the uncertainty (the location)
2) How well the uncertainty can be quantified (the level), and
3) Whether or not the uncertainty is believed to be caused by natural variance or a lack of information (Nature)

 . Additionally, since many papers will go on to define uncertainty as a probability or a density in their method, the actual organisation of "uncertainty" is irrelevant to their results since a probability distribution is

or where it exists in the broader scope of uncertainty communication. If a taxonomy is presented it is rarely considered when discussing differences in calculating, conceptualising,  The definition is also rarely dicussed beyond the introudction of a paper. to its place in the broader scope of of these definitions are presented with little justification as to why

Type A evaluation of standard uncertainty, and the other which is based on scientific judgement (such as experience and specifications), called the Type B evaluation of standard uncertainty.

It can be seen that while these taxonomies are both attempting to organise similar aspects of uncertainty, they have little overlap. The reason for this is relatively simple. Uncertainty is often only defined in so far as it is needed.

Ultimately these taxonomies end up alienating those that actually work with uncertainty by being too narrow. Uncertainty is a broad and complicated topic and while oversimplified taxonomies might help us hold the concept of uncertainty in our head, they do little to help with the practical question of how to quantify, visualise, or communicate uncertainty.

- mention that utypo taxonomy contains all these taxonomies and can be easily built upon without making a new taxonomy.

@@

This way of conceptualising and organising uncertainty is not a fringe idea and is backed up by several related ideas in statistics. 
- also makes it easier for people to communicate exactly what they need visualisation tools for. Look at geography papers and how their specific uncertainty considerations have been ignored for years.
- can organise other "uncertainties" according to this framework
- it also gives us rules for combining uncertainty since statistical uncertainty can be combined with staitstical uncertainty, and outcome uncertainty can be combined with outcome uncertainty.
Additionally, this definition of uncertainty is helpful in understanding *how* uncertainty should be communicated and the consideration we should make.

*Notes from other section*
- The usually uncertainty is only considered at the final stage but it needs to be considered through all stages. The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
- @Meng2021 highlights that data needs to be treated as an output from another process rahter than an value neutral input.
- If the distinction between noise and uncertainty is largely task dependent, considerations of quantifying and expressing uncertainty should be considered for the entire duraction of a project; another concept that has been independently identified by several authors [@Kinkeldey2014; @Refsgaard2007]
- The most encompassing definition of uncertainty I have seen comes from @utypo who define uncertainty as **"any deviation from the unachievable ideal of completely deterministic knowledge of the relevant system"**. This definition encapsulates many concepts adjacent to randomness such as predictions, probability mass distributions (PMF), estimate error, and any data that is not a set of deterministic outcomes. This is the definition of uncertainty I will use for the rest of this report.
- While quantification of the uncertainty that arrises from statistical philosophy and data provenance are beyond the scope of this paper, how this uncertainty is communicated is not. The ontological process discussed by @Otsuka2023 where information is tossed as we boil our information down into statistical objects requires us to map our statistical objects back to real world entities. This means the noise introduced and signal lost in earlier stages of the data analysis need to be identified, quantified and expressed in terms of real world outcomes. Visualisation is often the final stage of larger data science project and the down stream effects of poor abstraction and problem at the lower levels become impossible to ignore [@Munzner2009]. 
- Approaches to communicating limited understanding include 1) explicit model uncertainty 2) Qualitative scales expressed as strength of evidence (e.g. high to low certainty of evidence) 3) Acknowledge limitations and their possible impact 4) Acknowledged ignorance @Spiegelhalter2017

@@

- A lot of taxonomies for uncertainty, simple is "known knowns, known unknowns, and unknown unknowns", ka (known, unknown, nad unknowable), (risk, uncertainty, ignorance). This also breaks down into the framework of aleatory uncertainty (unavoidable randomness), epistemic uncertatiny (uncertainty about structure, still quantified), ontological uncertainty (uncertain about entire modelling process). @Spiegelhalter2017
- Can discuss each uncertainty in isolation using distributions, ranges, and lists of alternatives, but the total cumulative uncertainty from every stage is more difficult (near impossible) @Spiegelhalter2017
# 3.General Attitude Towards Uncertainty + Its communication
I dont think this is worth a section but it is mentioned a lot in .
- the role of unertainty visualisation is to communicate accurate information
- problem of communicating to a group thing (avoid communicating inaccurate information)
- how people feel about uncertainty is irrelevant
  - natural disasters make people uncomfortable but to include it in a conversation about whether or not we should express it is rediculous.
- Numerical risks: Use absolute risks (but relative risk when catastrophic event), for single unique event use percent chance, express change as a proportion, frequency, or percentage (and be clear about reference class), avoid framing bias by showing risk with and without the outcome, keep the denominator fixed when making comparisons, explicit about time interval, For more knowledgeable audiences consider providing epistemic uncertainty and confidence in analysis @Spiegelhalter2017

@@

- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
# 5 Great Examples
- best things you have seen and why they are the best
- can't have a best practices because you need experimental results and my point is that the experiments aren't testing the right things
- VSUP
- the map interactive map that susuan showed me
- the pixel map
- The new york times wealthy people that end up in different economics classes, white vs black people
- the animated surface classifications from that satelite images experimental paper
- the climate change scenario uncertainty

### 2024-01-23 writingdumpster.qmd
started doing summary of general ideas (no fine details)

---
title: "Writing Trash Bin"
author: "Harriet Mason"
format: html
---
# Why this exists
I have an information hoarding problem, this helps me throw out sections of the literature review that are not necessarily relevant without actually throwing it out. Knowing that I can come back to this at any point and pull information back, makes me more comfortable with removing it from the report.

# context section outtakes
Given that uncertainty estimation and communication seem to be largely task depenent, what questions should we be testing? This leads to several unanswered questions. Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models? Is it to supress signal that could be the result of random variance? Is it to find the best way to convey the mass function of a random variable? A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.
This conceptual way of thinking about uncertainty has been rattling around in the different areas of data science for a while.
For example, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.
@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.  @Munzner2009 created a nested model for visualisation that highlighted how pitfalls in upstream levels, such as problem characterisation and data abstraction, can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.

>>>>

---
title: "Writing Trash Bin"
author: "Harriet Mason"
format: html
---
# Why this exists
I have an information hoarding problem, this helps me throw out sections of the literature review that are not necessarily relevant without actually throwing it out. Knowing that I can come back to this at any point and pull information back, makes me more comfortable with removing it from the report.
# Definition outtakes
## Random
- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
-  @Spiegelhalter2017 defines risk as risk known probabilities and uncertainty as estimated probabilities, therefore uncertainty is what we have when you cannot accurately quantify risk.

## Taxonomies
To avoid alienating lay people with confusing complexity, many of these taxonomies are overly simplistic. Most taxonomies seek to organise uncertainty based off where the uncertainty is coming from. For example, @Spiegelhalter2017 organises uncertainty into three categories: 
  - Aleatory uncertainty: inevitable unpredictability due to unforeseeable factors 
  - Epistemic uncertainty: uncertainty about the structure and parameters of statistical models
  - Ontological uncertainty: uncertainty about the entire modelling process as a description of reality.
These three categories can be considered to be uncertainty when deciding which model we should use (ontological), uncertainty surrounding parameter estimates (epistemic), and uncertainty surrounding a forecast (aleatory), however it is not hard to find another, very different, example of a taxonomy that organises uncertainty based on what is seen to be causing it. @Gustafson2019 organises uncertainty into the four groups:
  - Deficient: Lack of knowledge
  – Technical: modelling approximations and measurement error
  – Scientific Unknown unknowns
  – Consensus: Disagreement among parties
However the additional comments about the inherent variability of aleatory uncertainty results in a bit of confusion as to if our cateogirisation should change if we decide to treat the parameters of a statistical model to be inherently random (as is done in Bayesian statistics) or if we beleive a forecast variability to be caused by a lack of knowledge.

He kept the taxonomy to three levels to maintain simplicity, however this organisation methods conflates amount of uncertainty present in an estimate with what is causing that uncertainty, and in doing so makes it hard for us to use this definition to better understand uncertainty within our own analysis.
# context section outtakes
Given that uncertainty estimation and communication seem to be largely task depenent, what questions should we be testing? This leads to several unanswered questions. Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models? Is it to supress signal that could be the result of random variance? Is it to find the best way to convey the mass function of a random variable? A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.
This conceptual way of thinking about uncertainty has been rattling around in the different areas of data science for a while.
For example, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.
@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.  @Munzner2009 created a nested model for visualisation that highlighted how pitfalls in upstream levels, such as problem characterisation and data abstraction, can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
